Railway: Adaptive Disk Storage for Interaction Graphs

### 2014-09-11 docs/edbt2015/text/introduction.tex
first cut of the introduction

Interaction graphs are temporal in nature, and more importantly, they are
append-only. This is in contrast to relationship graphs, which are updated via
insertion and deletion operations. An example of a relationship graph is a
social network capturing the follower-followee relationship among users.
Examples of interactions graphs include CDR graphs capturing calls between
telco customers or mention graphs capturing interactions between users of a
micro-blogging service. The append-only nature of the interaction graphs make
storing them on disk a necessity. Furthermore, the analysis of this historical
interaction data form an important aspect of the analytical landscape.

The ability to efficiently support historical analysis over interaction graphs
require effective solutions for the problem of \emph{data layout} on disk.
Most graph algorithms are characterized by locality of
access~\cite{localityLayout}, which has been taken advantage of co-locating
edges in close proximity within the same disk blocks~\cite{gstoreThesis}. This
ensures that once a disk block is loaded from disk into the main memory
buffers, several edges can be used. The locality of reference is a result of
the traversal-based nature of most of the graph algorithms.

In interaction graphs, the locality of access is even more pronounced. First, 
the analysis to be performed on the interactions can be restricted to a
temporal view of the graph, such as finding the influential users over a given
week of interactions. This means that edges that are temporally close are to
be accessed together. Second, traversals are again key to many algorithms,
such as connected components, clustering coefficient, PageRank, etc. This
means that edges that are close by in terms of the path between their incident
vertices as well as their timestamps are to be located together with the same
blocks.

In our earlier work~\cite{gedik14}, we introduced an interaction graph database
that works on this principle of access locality. It proposed a disk
organization that consists of as set of blocks, each containing a list of
\emph{temporal neighbor lists}. A temporal neighborlist contains a head vertex
and a set of incident edges within a given time range. The layout optimizer
aims at bringing together temporal neighborlists that are (i) close in terms
of their temporal ranges, (ii) have many edges between them, and (iii) have
few edges going into temporal neighborlists outside the block.

Many real world graph databases contain attributes. In the case of interaction
graphs, the attributes can be considered as properties associated with the
edges. Attributes can be stored in two ways, either separately (e.g., in a
relational table), or locally with the temporal neighbor lists.  If they are
stored separately, then the graph database cannot take advantage of locality
optimizations performed for block organization. The database must go back and
forth between the disk blocks to access the edge attributes.  On the other
hand, if attributes are stored locally in the disk blocks containing the graph
structure, then there can be significant overhead due to disk I/O if only a
few attributes are needed to answer a query.

@@

It is widely recognized that query workload and disk layout have a significant
impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
table-based, relational databases, this fact has led database designers to
develop alternative approaches for storage layout: row-oriented
storage~\cite{rowOrg} is more efficient when queries access many attributes
from a small number of records, and column-oriented storage~\cite{colOrg} is
more efficient when queries access a small number of attributes from many
records~\cite{stonebraker05}.  Unfortunately, although interaction graph
databases, like relational databases, are the target of diverse query
workloads, there is no clear correspondence to a row-oriented or
column-oriented storage layout.

@@

There a number of challenges in achieving an effective adaptive layout. First,
we need to find the partitioning of attributes that optimizes the query I/O.
To address this, we model the problems of overlapping and non-overlapping
attribute partitioning as mixed-integer linear programs (ILPs), and provide
optimal solutions that minimize the I/O cost. Second, the query workload, and
thus the attribute access pattern can change over time. For this purpose, our
railway layout supports customization of the attribute partitioning of
sub-blocks on a per-block basis. Third, such flexibility necessitates online
configuration of attribute partitioning as the query workload evolves, which
in turn requires fast algorithms for performing the attribute partitioning.
For this purpose, we develop greedy heuristic algorithms for both overlapping
and non-overlapping partitioning scenarios. These algorithms can scale to
larger number of attributes, yet provide close to optimal query I/O
performance. Finally, the railroad layout tradeoffs storage space to gain
improved query I/O performance. The storage overhead is more pronounced for
the case of overlapping partitioning. To address this, we limit the amount of
storage  overhead that can be tolerated with an upper limit, and integrate
this limit to both our ILP formulations, as well as our greedy heuristics.

>>>>

Interaction graphs are temporal in nature, and more importantly, they are
append-only. This is in contrast to relationship graphs, which are updated via
insertion and deletion operations. An example of a relationship graph is a
social network capturing the follower-followee relationship among users.
Examples of interactions graphs include CDR graphs capturing calls between
telco customers or mention graphs capturing interactions between users of a
micro-blogging service. The append-only nature of the interaction graphs make
storing them on disk a necessity. Furthermore, the analysis of this historical
interaction data form an important part of the analytical landscape.

The ability to efficiently support historical analysis over interaction graphs
require effective solutions for the problem of \emph{data layout} on disk.
Most graph algorithms are characterized by locality of
access~\cite{localityLayout}, which is a direct result of the traversal-based
nature of most of the graph algorithms. This is often taken advantage of by
co-locating edges in close proximity within the same disk
blocks~\cite{gstoreThesis}. This way, once a disk block is loaded into main
memory buffers, several edges from it can be used for processing, reducing the
disk I/O.

In interaction graphs, the locality of access is even more pronounced. First, 
the analysis to be performed on the interactions can be restricted to a
temporal view of the graph, such as finding the influential users over a given
week of interactions. This means that edges that are temporally close are to
be accessed together. Second, traversals are again key to many graph
algorithms, such as connected components, clustering coefficient, PageRank,
etc. This means that edges that are close by in terms of the path between
their incident vertices as well as their timestamps should be located together
with the same blocks.  In our earlier work~\cite{gedik14}, we introduced an
interaction graph database that works on this principle of access locality. It
uses a disk organization that consists of as set of blocks, each containing a
list of \emph{temporal neighbor lists}. A temporal neighbor list contains a
head vertex and a set of incident edges within a time range. The layout
optimizer aims at bringing together, into the same disk block, temporal
neighbor lists that are ($i$) close in terms of their temporal ranges, ($ii$)
have many edges between them, and ($iii$) have few edges going into temporal
neighbor lists outside the block.

Many real world graph databases contain attributes. In the case of interaction
graphs, the attributes can be considered as properties associated with the
edges representing the interactions. Attributes can be stored in two ways,
either separately (e.g., in a relational table), or locally with the temporal
neighbor lists.  If they are stored separately, then the graph database cannot
take full advantage of locality optimizations performed for block
organization. The database must go back and forth between the disk blocks to
access the edge attributes.  On the other hand, if attributes are stored
locally in the disk blocks containing the graph structure, then there can be
significant overhead due to disk I/O if only a few attributes are needed to
answer a query.

@@

It is widely recognized that query workload and disk layout have a significant
impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
table-based relational databases, this fact has led database designers to
develop alternative approaches for storage layout: row-oriented
storage~\cite{rowOrg} is more efficient when queries access many attributes
from a small number of records, and column-oriented storage~\cite{colOrg} is
more efficient when queries access a small number of attributes from many
records~\cite{stonebraker05}.  Unfortunately, although interaction graph
databases, like relational databases, are the target of diverse query
workloads, there is no clear correspondence to a row-oriented or
column-oriented storage layout.

@@

There a number of challenges in achieving an effective adaptive layout. First,
we need to find the partitioning of attributes that minimized the query I/O.
To address this, we model the problems of overlapping and non-overlapping
attribute partitioning as mixed-integer linear programs (ILPs), and provide
optimal solutions that minimize the query I/O cost. Second, the query
workload, and thus the attribute access pattern can change over time. For this
purpose, our railway layout supports customization of the attribute
partitioning of sub-blocks on a per-block basis. Third, such flexibility
necessitates online configuration of attribute partitioning as the query
workload evolves, which in turn requires fast algorithms for performing the
attribute partitioning. For this purpose, we develop greedy heuristic
algorithms for both overlapping and non-overlapping partitioning scenarios.
These algorithms can scale to larger number of attributes, yet provide close
to optimal query I/O performance. Finally, the railroad layout trades off
storage space to gain improved query I/O performance. The storage overhead is
more pronounced for the case of overlapping partitioning. To address this, we
limit the amount of storage  overhead that can be tolerated, and integrate
this limit to both our ILP formulations, as well as our greedy heuristics.
